Introducing statistics with \\texttt{intRo}

### 2014-12-12 student-paper-competition-2014/paper.Rmd
Edits from reading through paper

\begin{abstract}
\texttt{intRo} is a modern web-based application for performing basic data analysis and statistical routines. Leveraging the power of \texttt{R} and \texttt{Shiny}, \texttt{intRo} implements common statistical functions in a powerful and extensible modular structure, while remaining simple enough for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages, such as \texttt{Excel} and \texttt{JMP}. \texttt{intRo} is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper we introduce the application and explore the design decisions underlying \texttt{intRo} as well as highlight some challenges and advantages of reactive programming.
\end{abstract}

@@

Other software has recently been written in an attempt to spur student interest in introductory statistics and R programming. DataCamp's [@datacamp] courses are user-friendly ways to learning basic \texttt{R} programming and data analysis techniques. \texttt{Swirl} [@swirl] is a similar interactive tool to make learning \texttt{R} more fun by learning \texttt{R} within \texttt{R} itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach \texttt{R} programming, rather than facilitate the learning of introductory statistics. Project MOSAIC's goal is to faciliate this learning, but using the package requires a knowledge of R programming that the introductory student may not have.

@@

 1. **Top Navigation** - The top navigation bar includes two sets of buttons. The left-aligned buttons are informational buttons. The first is a link to \texttt{intRo}. The second is a link to this documentation page. The third is a link to the GitHub repository where the code for \texttt{intRo} is housed. The final button is a link to our websites, which contain contact information if there are any questions or comments. The right-aligned buttons are \texttt{intRo} utilities. The first is a button to toggle the visibility of the code panel (5). The last is a button to print the active module results, and the associated code (if visible).
 2. **Side Navigation** - The side navigation panel includes the modules that perform a specific data analysis task.
 3. **Options Panel** - The options panel includes module-specific options which the user can use to customize their results.
 4. **Results Panel** - The results pane displays the result of the selected module and options.
 5. **Code Panel** - The code panel displays the \texttt{R} code used to generate the results from the student's \texttt{intRo} session. The code panel is shown by default to facilitate an interest in coding, but can be hidden by clicking the code toggle button in the Top Navigation bar.

@@

In addition to making \texttt{intRo} intuitive to use, we've also created a documentation website that is consistent with the interface of the application. The documentation is available at http://gammarama.github.io/intRo/ and hyperlinked within the application itself. This documentation covers all default modules and makes learning to use \texttt{intRo} even more painless.

## Exciting
In general, introduction to statistics is a class undergraduate students dread. It's hard, the language is unfamiliar, and the software is hard to use. With \texttt{intRo}, we hope to flip this paradigm. \texttt{intRo} is a fun, easy to use tool that's available on the web, and is also usable on mobile devices. As a web-based application, this tool is immediately more familiar to students than a desktop application. With interactive plots from \texttt{ggvis} [@ggvis], \texttt{intRo} brings a `wow'-factor to learning introductory statistics.

@@

## Extensible
\label{sec:extensible}
On the front end, user interaction with \texttt{intRo} is split into bitesize chunks that we call modules. This is true of the back end as well. Each module is a self contained set of \texttt{R} code that is dynamically added to the application at run time. What this means is that \texttt{intRo} is easily extended by the addition of modules within the framework underlying the application. This extensibility allows for \texttt{intRo} to be customized for an individual classroom setting, tailored to the needs of a particular course.

As an example of possible customization, an instructor may wish to introduce hypothesis testing from a randomization framework rather than distributional. In this case, a non-parametric test could be added easily as a module to be used instead of the T test. This addition would be fairly simple to complete. The user would fork \texttt{intRo} from http://github.com/gammarama/intRo and begin by adding six types of files to a folder within a module folder: `helper.R`, `observe.R`, `output.R`, `reactive.R`, `static.R`, and `ui.R`. Each of these files is written to comply with the reactive nature available in a \texttt{Shiny} application. The `ui.R` file contains all the user interface elements within the new module's tab. This will include any options in the options panel as well as results. The `helper.R` and `static.R` files refer to static functions and global variable that may be used within the module. The file `output.R` contains the reactive elements that tie output to the user interface within \texttt{Shiny} and `observe.R` and `reactive.R` contain elements within an `observer` function or `reactive` function within the \texttt{Shiny} framework. By placing each of these files within the `modules/nonparametric_test` folder and adding a single line to a navigation file indicating under which module grouping the non-parametric test module should go, a new module is available within \texttt{intRo} upon running the application.

@@

\begin{table*}[ht!]
\centering
\begin{tabular}{|c|c|c|}
\hline
Plot Type&X Variable Type Allowed&Y Variable Type Allowed\\
\hline
Histogram&Numeric&N/A\\
Normal Quantile Plot&Numeric&N/A\\
Scatterplot&Numeric&Numeric\\
Line Chart&Numeric&Numeric\\
Boxplot&Categorical&Numeric\\
Bar Chart&Categorical&Numeric\\
Pareto Chart&Categorical&Numeric\\
Mosaic Plot&Categorical&Categorical\\
\hline
\end{tabular}
\caption{The plot and variable types allowed in \texttt{intRo}'s graphical module. For example, when the selected plottype is histogram, the interface displays only a single variable selection box (x), and the variables selectable in this box are only numeric. If a boxplot is selected, the x variable must be categorical and the y variable must be numeric, and the list of choices updates automatically to reflect this.}
\label{tab:plottypes}
\end{table*}
Part two of this text discusses probability and sampling distributions, which is outside the current realm of \texttt{intRo}, but part three focuses on inferential statistics. The inferential statistics include confidence intervals, hypothesis testing, and comparing two variables, all supported natively in \texttt{intRo}. Finally, part four of the text discusses the association between variables, including regression, functions that \texttt{intRo} performs with ease. Some of the more advanced topics of this section are currently unsupported by \texttt{intRo}, including multiple regression and nonparametric statistics.

@@

\texttt{intRo} leverages reactive programming techniques as data is manipulated by the user, this carries throughout the entire session. In a typical \texttt{Shiny} app, users interact with inputs, which in turn yield different results. Within \texttt{intRo}, the users are able to interact and manipulate the data underlying the entire application. This posed many challenges in the creation of \texttt{intRo}, namely timely save points according to the user's timeline, and reactive updating of variable lists tied to inputs across the entire application. Because the user may experiment with different configurations or select different variables, we did not want to store all actions taken in the intRo session. Rather, each module includes a button allowing the user to explicitly store the output visible in the results panel into the R script. This way, output is only stored when the user is satisfied, and the resulting output is not cluttered with unnecessary information.

In the creation of \texttt{intRo} we walked a fine line between giving the user flexibility and having realistic usability. At the same time, \texttt{intRo} was created as a consumer of another package, \texttt{Shiny}, in which we as developers were the beneficiaries of another team of developers' decision to balance flexibility and usability. For a tangible example, consider the graphical summaries module. We only allow variables of a type consistent with the selected plot to be displayed. This is a conscious decision that limits an \texttt{intRo} user's flexibility, while maximizing the usability (by minimizing crashes) of the application. On the flip side of this, \texttt{Shiny} allows much higher flexibility. For instance, the entire application (including user interface) is created dynamically upon load, based on the modules currently housed within \texttt{intRo}, as detailed in section \ref{sec:extensible}. However, \texttt{Shiny} does have limits in its flexibility based on the designers decisions for usability. One current example the slider element. This element allows for fixed width steps from its minimum to its maximum. The JavaScript library being utilized in \texttt{Shiny} allows for arbitrary function calls to to generate these steps, however they must be written in plain JavaScript. This is an example of a decision made by the developers of \texttt{Shiny} to limit functionality in favor of usability of their package.

>>>>

\begin{abstract}
\texttt{intRo} is a modern web-based application for performing basic data analysis and statistical routines. Leveraging the power of \texttt{R} and \texttt{Shiny}, \texttt{intRo} implements common statistical functions in a powerful and extensible modular structure, while remaining simple enough for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages, such as \texttt{Excel} and \texttt{JMP}. \texttt{intRo} is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper, we introduce the application and explore the design decisions underlying \texttt{intRo}, as well as highlight some challenges and advantages of reactive programming.
\end{abstract}

@@

Other software has recently been written in an attempt to spur student interest in introductory statistics and R programming. DataCamp's [@datacamp] courses are user-friendly ways to learning basic \texttt{R} programming and data analysis techniques. \texttt{Swirl} [@swirl] is a similar interactive tool to make learning \texttt{R} more fun by learning it within \texttt{R} itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach \texttt{R} programming, rather than facilitate the learning of introductory statistics. Project MOSAIC's goal is to faciliate this learning, but using the package requires a knowledge of R programming that the introductory student may not have.

@@

 1. **Top Navigation** - The top navigation bar includes two sets of buttons. The left-aligned buttons are informational buttons. The first is a link to \texttt{intRo}. The second is a link to this documentation page. The third is a link to the GitHub repository where the code for \texttt{intRo} is housed. The final button is a link to our websites, which contain contact information if there are any questions or comments. The right-aligned buttons are \texttt{intRo} utilities. The first is a button to toggle the visibility of the code panel (5). The last is a button to print the stored module results, and the associated code (if visible).
 2. **Side Navigation** - The side navigation panel includes the modules that perform a specific data analysis task.
 3. **Options Panel** - The options panel includes module-specific options which the user can use to customize their results.
 4. **Results Panel** - The results pane displays the result of the selected module and options.
 5. **Code Panel** - The code panel displays the \texttt{R} code used to generate the results from the student's \texttt{intRo} session. The code panel is shown by default to facilitate an interest in coding, but can be hidden by clicking the code toggle button in the Top Navigation bar.

@@

In addition to making \texttt{intRo} intuitive to use, we've also created a documentation website that is consistent with the interface of the application. The documentation is available at http://gammarama.github.io/intRo and hyperlinked within the application itself. This documentation covers all default modules and makes learning to use \texttt{intRo} even more painless.

## Exciting
In general, introduction to statistics is a class undergraduate students dread. It's hard, the language is unfamiliar, and the software is difficult to use. With \texttt{intRo}, we hope to flip this paradigm. \texttt{intRo} is a fun, easy to use tool that's available on the web, and is also usable on mobile devices. As a web-based application, this tool is immediately more familiar to students than a desktop application. With interactive plots from \texttt{ggvis} [@ggvis], \texttt{intRo} brings a `wow'-factor to learning introductory statistics.

@@

## Extensible
\label{sec:extensible}
On the front end, user interaction with \texttt{intRo} is split into bitesize chunks that we call modules. This is true of the back end as well. Each module is a self contained set of \texttt{R} code that is dynamically added to the application at run time. What this means is that \texttt{intRo} can be easily extended by the addition of modules within the framework underlying the application. This extensibility allows for \texttt{intRo} to be customized for an individual classroom setting, tailored to the needs of a particular course.

As an example of possible customization, an instructor may wish to introduce hypothesis testing from a nonparametric framework rather than distributional. In this case, a wilcoxon rank sum test could be added easily as a module to be used instead of the T test. This addition would be fairly simple to complete. The user would fork \texttt{intRo} from http://github.com/gammarama/intRo and begin by adding six types of files to a folder within a module folder: `helper.R`, `observe.R`, `output.R`, `reactive.R`, `static.R`, and `ui.R`. Each of these files is written to comply with the reactive nature available in a \texttt{Shiny} application. The `ui.R` file contains all the user interface elements within the new module's tab. This will include any options in the options panel as well as results. The `helper.R` and `static.R` files refer to static functions and global variable that may be used within the module. The file `output.R` contains the reactive elements that tie output to the user interface within \texttt{Shiny} and `observe.R` and `reactive.R` contain elements within an `observer` function or `reactive` function within the \texttt{Shiny} framework. By placing each of these files within the `modules/nonparametric_test` folder and adding a single line to a navigation file indicating under which module grouping the non-parametric test module should go, a new module is available within \texttt{intRo} upon running the application.

@@

\begin{table*}[ht!]
\centering
\begin{tabular}{|c|c|c|}
\hline
Plot Type&X Variable Type Allowed&Y Variable Type Allowed\\
\hline
Histogram&Numeric&N/A\\
Normal Quantile Plot&Numeric&N/A\\
Scatterplot&Numeric&Numeric\\
Line Chart&Numeric&Numeric\\
Boxplot&Categorical&Numeric\\
Bar Chart&Categorical&Numeric\\
Pareto Chart&Categorical&Numeric\\
Mosaic Plot&Categorical&Categorical\\
\hline
\end{tabular}
\caption{The plot and variable types allowed in \texttt{intRo}'s graphical module. For example, when the selected plottype is histogram, the interface displays only a single variable selection box (x), and the variables selectable in this box are only numeric. If a boxplot is selected, the x variable must be categorical and the y variable must be numeric, and the list of choices updates automatically to reflect this.}
\label{tab:plottypes}
\end{table*}

Part two of this text discusses probability and sampling distributions, which is outside the current realm of \texttt{intRo}, but part three focuses on inferential statistics. The inferential statistics include confidence intervals, hypothesis testing, and comparing two variables, all supported natively in \texttt{intRo}. Finally, part four of the text discusses the association between variables, including regression, functions that \texttt{intRo} performs with ease. Some of the more advanced topics of this section are currently unsupported by \texttt{intRo}, including multiple regression.

@@

\texttt{intRo} leverages reactive programming techniques as data is manipulated by the user, and this is carried throughout the entire session. In a typical \texttt{Shiny} app, users interact with inputs, which in turn yield different results. Within \texttt{intRo}, the users are able to interact and manipulate the data underlying the entire application. This posed many challenges in the creation of \texttt{intRo}, namely timely save points according to the user's workflow, and reactive updating of variable lists tied to inputs across the entire application. Because the user may experiment with different configurations or select different variables, we did not want to store all actions taken in the intRo session. Rather, each module includes a button allowing the user to explicitly store the output visible in the results panel into the R script. This way, output is only stored when the user is satisfied, and the resulting output is not cluttered with unnecessary information.

In the creation of \texttt{intRo} we walked a fine line between giving the user flexibility and having realistic usability. At the same time, \texttt{intRo} was created as a consumer of another package, \texttt{Shiny}, in which we as developers were the beneficiaries of another team of developers' decision to balance flexibility and usability. For a tangible example, consider the graphical summaries module. We only allow variables of a type consistent with the selected plot to be displayed. This is a conscious decision that limits an \texttt{intRo} user's flexibility, while maximizing the usability (by minimizing crashes) of the application. On the flip side of this, \texttt{Shiny} allows much higher flexibility. For instance, the entire application (including user interface) is created dynamically upon load, based on the modules currently housed within \texttt{intRo}, as detailed in section \ref{sec:extensible}. However, \texttt{Shiny} does have limits on its flexibility based on the designers decisions for usability. One current example the slider element. This element allows for fixed width steps from its minimum to its maximum. The JavaScript library being utilized in \texttt{Shiny} allows for arbitrary function calls to to generate these steps, however they must be written in plain JavaScript. This is an example of a decision made by the developers of \texttt{Shiny} to limit functionality in favor of usability of their package.
